---
title: "Outbreak Detection"
output: html_document
date: "2024-11-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


## Expected learning outcomes

By the end of this session, participants should be able to:

- understand the basics principles of outbreak detection including creating and expectation, threshold and alarm

- implement & apply different outbreak detection methods to timeseries with different characteristics and understand advantages and disadvantages of different methods

- get to know the SignalDetectionTool and be able to use functions for outbreak detection from the tool as well as the shiny app

## Preparations

Required source code to load.
In case participants could not install the package SignalDetectionTool they can source the code source_signaldetectiontool.R which has all functions required for Section 4.2. . 

```{r, include = TRUE, source-Rscript}
source("source_outbreak_det.R")
# source("source_signaldetectiontool.R")
```

Required packages

```{r, load-packages, include = TRUE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(surveillance)
library(tidyr)
library(SignalDetectionTool)
```

# Section 1: Implement your first outbreak detection algorithm

We will now import the first dataset to analyse.

```{r, load-dataset}
ts0 <- readRDS("data/sim0.Rds")
```

## Task 1.1 
Visualise the first timeseries ts0 using the function `plot_ts()` which is provided in the source_outbreak_det.R script.
What do you observe? What characteristics does this timeseries have? Do you see any potential outbreaks?

```{r, vis-dataset-task-1.1}
plot_ts(ts0)
```
No specific patterns (no obvious trend, no seasonality) thus stationary timeseries. Some potential outbreaks at the end of the timeseries and in spring 2019.

## Task 1.2

We want to apply a simple outbreak detection method to the last timepoint of this timeseries.
The baseline/expectation should be the mean of all observations before current time point. For calculating a threshold we are going to use the 95% quantile of the Poisson distribution. As later in this practical we want to add other algorithms we have an additional parameter called method. 
Please fill all the spots with ____ to make this function work. 

```{r, outbreak-detection-algo-task-1.2}
detect_outbreaks <- function(ts, method) {
  length_ts <- nrow(ts)
  # split the dataset
  # datapoints we use for model fitting, i.e. we build our expecation on
  fit_data <- ts[1:(length_ts - 1), ]
  # the last datapoint we want to do signal detection for
  pred_data <- ts[length_ts, ]

  # fit the model
  if (method == "average") {
    model <- glm(formula = cases ~ 1, data = fit_data, family = poisson())
  }
  # we generate an expectation and threshold for visualisation purposes for the data used for fitting
  # model$fitted.values give the expected values based on the model already transformed using exp() (see documentation)
  fit_data <- fit_data %>% mutate(
    expectation = model$fitted.values,
    threshold = qpois(0.95, expectation),
    alarms = NA_real_
  )

  # extend the fitted model (make predictions) for the last data point which we want to do signal detection for
  pred <- predict.glm(model,
    newdata = pred_data
  )

  pred_data <- pred_data %>%
    mutate(
      expectation = exp(pred),
      threshold = qpois(0.95, expectation),
      alarms = pred_data$cases > threshold
    )

  # combine the two datasets back together
  bind_rows(
    fit_data,
    pred_data
  )
}
```

## Task 1.3

Now apply the `detect_outbreaks()` function for outbreak detection to the timeseries using the method "average" and plot the result using `plot_ts()` with the additional argument `with_signals = TRUE`. Does what you observe match your expecations?

```{r,apply-detect-outbreaks-task-1.3}
signals_average <- detect_outbreaks(ts0, method = "average")
plot_ts(signals_average, with_signals = TRUE)
```
Observations:
We see a constant average as computed expectation. The last timepoint is marked as outbreak which is expected as it was a peak compared to what was observed before in the timeseries. 

# Section 2:  Apply the outbreak detection to multiple weeks

We want to extend the algorithm so that it can not only be applied to the most recent week but multiple past weeks. For this we implement a function which consists of for-loop around our function `detect_outbreaks()`. 

## Task 2.1
Please fill the ____ in this function. 
```{r, outbreak-detection-multiple-weeks-task-2.1}
detect_outbreaks_multiple_weeks <- function(ts, method, number_of_weeks = 4) {
  signal_results <- data.frame()
  len_ts <- nrow(ts)

  for (i in seq(0, number_of_weeks - 1, 1)) {
    ts <- ts[1:(len_ts - i), ]

    ts_with_signals <- detect_outbreaks(ts, method = method)

    signals <- ts_with_signals %>%
      tail(1)
    signal_results <- bind_rows(signal_results, signals)

    if (i == number_of_weeks - 1) {
      # want to get the modeled expectation and threshold for the data used for fitting as well
      # take the results from the timeseries with the earliest timepoint where signals are generated for
      ts_without_signal_weeks <- ts_with_signals[1:(nrow(ts_with_signals) - 1), ]
    }
  }
  # put timeseries and signal results together in one df
  ts_without_signal_weeks %>%
    bind_rows(signal_results) %>%
    arrange(date)
}
```

## Task 2.2

Apply the `detect_outbreaks_multiple_weeks()` to the ts0 using the method "average" for the most recent 4 weeks and visualise the result.

```{r, average-multiple-weeks-task-2.2}
ts0_signals_4_weeks <- detect_outbreaks_multiple_weeks(ts0, method = "average")
plot_ts(ts0_signals_4_weeks, with_signals = TRUE)
```

# Section 3: Improve the first simple outbreak detection algorithm

We will split into groups now and each group obtains a different timeseries. Please discuss within your groups and work together. At the end one person of each group will give a short (max 5 min) presentation of the results to everyone in the plenum. Please only work on the parts of this .rmd which belongs to your group.

## Group 1: 

We now load another timeseries and visualise it. 

```{r,group-1-timeseries-vis}
ts1 <- readRDS("data/sim1.Rds")
plot_ts(ts1)
```

### Task 3.1
How would you describe this timeseries? What characteristics do you see?

Increasing timetrend and small outbreak at the end.

### Task 3.2
Apply the simple average outbreak detection algorithm to this timeseries for the last 4 weeks and plot the results. 

```{r,group-1-average-application}
ts1_with_signals_average <- detect_outbreaks_multiple_weeks(ts1, method = "average")
plot_ts(ts1_with_signals_average, with_signals = TRUE)
```

### Task 3.3
What do you observe? Do you think this algorithm is suited for this timeseries? 

No it is not suited as it is not modeling the timetrend. We would get a lot of false alarms just because of the increasing case numbers because of the timetrend.

### Task 3.4
What are your ideas to improve this algorithm? Discuss in the group.

Add a timetrend term to the regression.

### Task 3.5
Implement your idea. If possible use the `detect_outbreaks()` function with a different method parameter. 
```{r,group-1-new-algo-task-3.5}
detect_outbreaks <- function(ts, method) {
  length_ts <- nrow(ts)
  # split the dataset
  # datapoints we use for model fitting, i.e. we build our expecation on
  fit_data <- ts[1:(length_ts - 1), ]
  # the last datapoint we want to do signal detection for
  pred_data <- ts[length_ts, ]

  # fit the model depending on the chosen method
  if (method == "average") {
    model <- glm(formula = cases ~ 1, data = fit_data, family = poisson())
  } else if (method == "timetrend") {
    model <- glm(formula = cases ~ 1 + time, data = fit_data, family = poisson())
  }
  # we generate an expectation and threshold for visualisation purposes for the data used for fitting
  # model$fitted.values give the expected values based on the model already transformed using exp() (see documentation)
  fit_data <- fit_data %>% mutate(
    expectation = model$fitted.values,
    threshold = qpois(0.95, expectation),
    alarms = NA_real_
  )

  # extend the fitted model (make predictions) for the last data point which we want to do signal detection for
  pred <- predict.glm(model,
    newdata = pred_data
  )

  pred_data <- pred_data %>%
    mutate(
      expectation = exp(pred),
      threshold = qpois(0.95, expectation),
      alarms = pred_data$cases > threshold
    )

  # combine the two datasets back together
  bind_rows(
    fit_data,
    pred_data
  )
}
```

### Task 3.6

Apply your new method to the timeseries for the last 4 weeks and visualise the result.

```{r,group-1-new-algo-application-task-3.6}
ts1_with_signals_timetrend <- detect_outbreaks_multiple_weeks(ts1, method = "timetrend")
plot_ts(ts1_with_signals_timetrend, with_signals = TRUE)
```
### Task 3.7
Did this improve the outbreak detection? 

Yes because now the timetrend is modeled as well and when using the average a lot of timepoints in the past would have been labeled as outbreaks which are not real outbreaks.

### Task 3.8
Prepare to give a max. 5min presentation of your findings to the plenum. 

## Group 2:
We now load another timeseries and visualise it.
```{r,group-2-timeseries-vis}
ts2 <- readRDS("data/sim2.Rds")
plot_ts(ts2)
```

### Task 3.1

How would you describe this timeseries? What characteristics do you see?

Falling timetrend. Outbreak at the end.

### Task 3.2
Apply the simple average outbreak detection algorithm to this timeseries for the last 4 weeks and plot the results. 
```{r,group-2-average-application-task-3.2}
ts2_with_signals_average <- detect_outbreaks_multiple_weeks(ts2, method = "average")
plot_ts(ts2_with_signals_average, with_signals = TRUE)
```

### Task 3.3
What do you observe? Do you think this algorithm is suited for this timeseries? 

No it is not suited as it is not modeling the timetrend. We are missing the outbreak at the end due to not modeling this.

### Task 3.4
What are your ideas to improve this algorithm? Discuss in the group.

Add a timetrend to the model.

### Task 3.5
Implement your idea. If possible use the `detect_outbreaks()` function with a different method parameter. 
```{r,group-2-new-algo-task-3.5}
detect_outbreaks <- function(ts, method) {
  length_ts <- nrow(ts)
  # split the dataset
  # datapoints we use for model fitting, i.e. we build our expecation on
  fit_data <- ts[1:(length_ts - 1), ]
  # the last datapoint we want to do signal detection for
  pred_data <- ts[length_ts, ]

  # fit the model depending on the chosen method
  if (method == "average") {
    model <- glm(formula = cases ~ 1, data = fit_data, family = poisson())
  } else if (method == "timetrend") {
    model <- glm(formula = cases ~ 1 + time, data = fit_data, family = poisson())
  }
  # we generate an expectation and threshold for visualisation purposes for the data used for fitting
  # model$fitted.values give the expected values based on the model already transformed using exp() (see documentation)
  fit_data <- fit_data %>% mutate(
    expectation = model$fitted.values,
    threshold = qpois(0.95, expectation),
    alarms = NA_real_
  )

  # extend the fitted model (make predictions) for the last data point which we want to do signal detection for
  pred <- predict.glm(model,
    newdata = pred_data
  )

  pred_data <- pred_data %>%
    mutate(
      expectation = exp(pred),
      threshold = qpois(0.95, expectation),
      alarms = pred_data$cases > threshold
    )

  # combine the two datasets back together
  bind_rows(
    fit_data,
    pred_data
  )
}
```

### Task 3.6

Apply your new method to the timeseries for the last 4 weeks and visualise the result.

```{r,group-2-new-algo-application-task-3.6}
ts2_with_signals_timetrend <- detect_outbreaks_multiple_weeks(ts2, method = "timetrend")
plot_ts(ts2_with_signals_timetrend, with_signals = TRUE)
```

### Task 3.7

Did this improve the outbreak detection? 

Yes the outbreak in the end is no longer missed because we are able to model the trend over time.

### Task 3.8
Prepare to give a max. 5min presentation of your results to the plenum. 

## Group 3:
We now load another timeseries and visualise it. 

```{r,group-3-timeseries-vis}
ts3 <- readRDS("data/sim3.Rds")
plot_ts(ts3)
```

### Task 3.1
How would you describe this timeseries? What characteristics do you see?

Seasonal structure. Outbreak at the end.

### Task 3.2

Apply the simple average outbreak detection algorithm to this timeseries for the last 4 weeks and plot the results. 

```{r,group-3-average-application}
ts3_with_signals_average <- detect_outbreaks_multiple_weeks(ts3, method = "average")
plot_ts(ts3_with_signals_average, with_signals = TRUE)
```

### Task 3.3
What do you observe? Do you think this algorithm is suited for this timeseries? 

The algorithm misses the seasonal structure and the outbreak at the end. No it is not suited.

### Task 3.4
What are your ideas to improve this algorithm? Discuss in the group.

Add seasonal terms to the regression using sine and cosine.

### Task 3.5
Implement your idea. If possible use the `detect_outbreaks()` function with a different method parameter. 


```{r,group-3-new-algo-task-3.5}
detect_outbreaks <- function(ts, method) {
  length_ts <- nrow(ts)
  # split the dataset
  # datapoints we use for model fitting, i.e. we build our expecation on
  fit_data <- ts[1:(length_ts - 1), ]
  # the last datapoint we want to do signal detection for
  pred_data <- ts[length_ts, ]

  # fit the model depending on the chosen method
  if (method == "average") {
    model <- glm(formula = cases ~ 1, data = fit_data, family = poisson())
  } # first solution using sin and cos
  else if (method == "seasonality") {
    model <- glm(formula = cases ~ 1 + sin + cos, data = fit_data, family = poisson())
  } # second solution using the month factor
  else if (method == "seasonality_month") {
    model <- glm(formula = cases ~ 1 + month, data = fit_data, family = poisson())
  }



  # we generate an expectation and threshold for visualisation purposes for the data used for fitting
  # model$fitted.values give the expected values based on the model already transformed using exp() (see documentation)
  fit_data <- fit_data %>% mutate(
    expectation = model$fitted.values,
    threshold = qpois(0.95, expectation),
    alarms = NA_real_
  )

  # extend the fitted model (make predictions) for the last data point which we want to do signal detection for
  pred <- predict.glm(model,
    newdata = pred_data
  )

  pred_data <- pred_data %>%
    mutate(
      expectation = exp(pred),
      threshold = qpois(0.95, expectation),
      alarms = pred_data$cases > threshold
    )

  # combine the two datasets back together
  bind_rows(
    fit_data,
    pred_data
  )
}
```

### Task 3.6
Apply your new method to the timeseries for the last 4 weeks and visualise the result.

Solution 1 using sin and cos
```{r,group-3-new-algo-application-task-3.6}
ts3_with_signals_season <- detect_outbreaks_multiple_weeks(ts3, method = "seasonality")
plot_ts(ts3_with_signals_season, with_signals = TRUE)
```

Solution 2 using month
```{r,group-3-new-algo-application-task-3.6-solution2}
ts3_with_signals_season_month <- detect_outbreaks_multiple_weeks(ts3, method = "seasonality_month")
plot_ts(ts3_with_signals_season_month, with_signals = TRUE)
```

### Task 3.7
Did this improve the outbreak detection? 

Yes the outbreak is now found which was not identified before because the model did not have seasonal terms and the peak was below the average.

### Task 3.8
Prepare to give a max. 5min presentation of your results to the plenum.

# Section 4: From linelist to signals
## Section 4.1: Using dplyr and the surveillance package

In the previous section you were always provided datasets which were already showing aggregated case counts by week and year. In the real world you will often start with a linelist of cases and need to aggregate it first to apply signal detection methods. In a previous practical of this module you already learned how to get from a linelist to an aggregated timeseries. We will shortly refresh this knowledge here and then further build and extend on it.

```{r,load-linelist}
# Load the dataset
data_linelist <- read.csv("data/linelist.csv")
```

### Task 4.1.1
Quickly familiarise yourself with the dataset. What variables does it contain and how does it look like? 

One line per case, we have dates such as date_report or date_onset where we could aggregate on. There are additional information for each case such as the pathogen the case had, age, sex, the region (county, state, ...).

We will aggregate the dataset to obtain the number of cases in each week of the year based on the reporting date `date_report`. 

```{r,aggregate-linelist}
input_aggr <- data_linelist %>%
  mutate(
    week = isoweek(date_report),
    year = isoyear(date_report)
  ) %>%
  summarise(cases = n(), .by = c(week, year))
```

### Task 4.1.2
Do you think we are done now? Do you have any thoughts what could be a problem here? Please stop for a second before continuing to read to think about this.

If there are some weeks where no cases were reported these weeks will be missing in the `input_aggr`. We will apply a function which is provided in the source script which adds these missing weeks with number of cases set to 0. 
```{r,fill-missing-weeks}
input_aggr_complete <- complete_weeks_aggr_data(input_aggr)
```

### Task 4.1.3
Is the `input_aggr_complete` longer than the original `input_aggr_complete`?

Now let's apply a outbreak detection method from the surveillance package. There are a lot of different R libraries containing signal detection methods and we want to shortly show you The surveillance package as it is a very popular package and still frequently used in this field.

Unfortunately the surveillance package uses its own data format to apply outbreak detection to a timeseries and thus first we need to transform our aggregated timeseries into this "sts" format.

```{r,transform-to-sts}
sts_timeseries <- sts(input_aggr_complete$cases,
  start = c(
    input_aggr_complete$year[1],
    input_aggr_complete$week[1]
  ),
  frequency = 52
)
```

### Task 4.1.4
Apply the EarsC algorithm `earsC()` to the timeseries and plot the results using the `plot()` function from the surveillance package. Use the default values of both functions and look up how to use the functions using ?earsC() or ?plot()

```{r,apply-ears-surveillance-task-4.1.4}
ears_results_surv <- earsC(sts_timeseries)
# also talk about that the output here is again in sts format
plot(ears_results_surv)
```

### Task 4.1.5
Have a look at the visualisation. Do you understand everything? If you have any questions about it discuss with your partner or ask one of the facilitators.

### Task 4.1.6
Let's say you are interested in more detailed analysis and want to stratify by age group. This means you want to apply signal detection methods to a timeseries for each age group. How would you need to change the steps above to do this? Discuss with a partner.

## Section 4.2: Usage of functions from the SignalDetectionTool

The R library SignalDetectionTool which was developed within the Joint Action United4Surveillance provides functions for you to simplify this process. We start again with the linelist of cases and apply the earsC algorithm to it but we use functions from the tool.

```{r,apply-ears-tool}
ears_results_tool <- data_linelist %>%
  preprocess_data() %>%
  get_signals(method = "ears", number_of_weeks = 6)
```

As you can see we do not need to do add weeks and year to the data_linelist, do not need to aggregate and transform to sts. All of this is done inside the functions `preprocess_data()` (adding of week and year and other preprocessing) and `get_signals()` (aggregating,transforming to sts if necessary,...) .

We can now plot the results using the plotting function from the SignalDetectionTool.
```{r,vis-ears-tool}
plot_time_series(ears_results_tool, interactive = TRUE)
```

Now let's do the desired stratified analysis by age group.

### Task 4.2.1
Have a look at the parameters of `get_signals()` and add the necessary parameter for getting signals for each age group in our dataset.

```{r,apply-ears-agegroup-tool-task-4.2.1}
ears_agegroup_results_tool <- data_linelist %>%
  preprocess_data() %>%
  get_signals(method = "ears", number_of_weeks = 6, stratification = "age_group")
```

### Task 4.2.3
Have a look at the signal results dataframe `ears_agegroup_results_tool` and visualise one age group of your choice using the `plot_time_series()` function.
```{r,vis-ears-agegroup-tool-task-4.2.3}
plot_time_series(ears_agegroup_results_tool %>% filter(stratum == "00-04"), interactive = TRUE)
```

## Section 4.3: Usage of the shiny app SignalDetectionTool
Now we want to go even one step further and use the shiny app SignalDetectionTool.
In general to start the tool you need to load the package SignalDetectionTool using `library(SignalDetectionTool)` (we already did this here) and then type `run_app()` into your Console in Rstudio.

### Task 4.3.1
Run the app and upload the linelist.csv inside the data folder and perform signal analysis stratified by age_group, sex and county.
