---
title: "Outbreak Detection"
output: html_document
date: "2024-11-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


## Expected learning outcomes

By the end of this session, participants should be able to:

- understand the basics principles of outbreak detection including creating and expectation, threshold and alarm

- implement & apply different outbreak detection methods to time series with different characteristics and understand advantages and disadvantages of different methods

- get to know the SignalDetectionTool and apply it to a real dataset to identify outbreaks

## Preparations

Required source code to load

```{r, include = TRUE, source-Rscript}
source("source_outbreak_det.R")
```

Required packages

```{r, load-packages, include = TRUE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(SignalDetectionTool)
```

# Section 1: Implement your first outbreak detection algorithm

We will now import the first data set to analyse.

```{r, load-dataset}
ts1 <- readRDS("data/sim3.Rds")
```

## Task 1.1 
Visualise the first time series ts1 using the function `plot_ts()` which is provided in the source_outbreak_det.R script.
What do you observe? What characteristics does this time series have? Do you see any potential outbreaks?

```{r, vis-dataset-task-1.1}
plot_ts(ts1)
```

## Task 1.2

We want to apply a simple outbreak detection method to the last time point of this time series.
The baseline/expectation of our outbreak detection method should be the mean of all observations before the current time point. Fill in the blanks to compute this baseline.

```{r, baseline-mean-1.2}

# split the data set
# data points we use for model fitting, i.e. we build our expectation on are all except the last one
fit_data <- ts1 %>% head(-1)
# the last data point we want to do signal detection for
pred_data <- ts1 %>% tail(1)

# fit the model
model <- glm(formula = _____, data = fit_data, family = poisson())

# we generate an expectation for visualisation purposes for the data used for fitting
# model$fitted.values give the expected values based on the model already transformed using exp() (see documentation)
fit_data <- fit_data %>% 
  mutate(
    expectation = model$fitted.values
    )

# based on the model make predictions for the last data point which we want to do signal detection for
pred <- predict.glm(model, newdata = pred_data)

pred_data <- pred_data %>%
  mutate(
    expectation = exp(pred)
    )

# combine the two data sets back together
data_with_baseline <- bind_rows(fit_data, pred_data)


```

## Task 1.3

Now plot the time series with the baseline using `plot_ts()`. Does what you observe match your expectations? Do you think this is a good fit?

```{r,plot_ts_with_baseline-1.3}


```


## Task 1.4

In the next step we calculate a threshold based on the 95% quantile of the Poisson distribution around the expected value of the last data point and add it to our pred_data. For visualisation purpose we also extend the threshold into the past.
Please fill all the spots with ____ to make the code work.

```{r, threshold-1.4}

pred_data <- pred_data %>%
  mutate(
    threshold = qpois(_____,____)
    )

fit_data <- fit_data %>%
  mutate(
    threshold = qpois(____,_____)
    )


# bind fit and pred data together

data_with_baseline_and_threshold <- bind_rows(fit_data, pred_data)
```

Please plot the time series with the expectation and the threshold. 

```{r,plot_baseline_threshold-1.4}


```


## Task 1.5

Lastly we want to generate signals. Signals are those case numbers that exceed the threshold. As we are performing outbreak detection for the last time point we only want to generate a signal for this time point. Please fill the blanks to add a signal to the pred_data.

```{r,signals-1.5}
pred_data <- pred_data %>%
  mutate(
    alarms = pred_data$cases > _____
    )

# again bind the results together
data_with_signals <- bind_rows(fit_data, pred_data)
```

## Task 1.6
Now we plot all elements we computed for the outbreak detection. Do you think this algorithm is suited for this timeseries? 
Do you have any suggestions for improvement? Discuss.

```{r,plot_ts_signals-1.5}


```

# Section 2: Improve the algorithm 

## Task 2.1

Improve the algorithm, have a look at the variables available in ts1. Please fill the gaps to implement the new algorithm. This time we directly compute the threshold and alarms as well in one step. 
```{r, improve_outbreak_det_method-2.1}
# first we make a fresh environment
rm(fit_data, pred_data, model, data_with_signals, envir = .GlobalEnv)

# split the data set
# data points we use for model fitting, i.e. we build our expectation on are all except the last one
fit_data <- ts1 %>% head(-1)
# the last data point we want to do signal detection for
pred_data <- ts1 %>% tail(1)

# fit the model
model <- glm(formula = cases ~ _____, data = fit_data, family = poisson())
# we generate an expectation for visualisation purposes for the data used for fitting
# model$fitted.values give the expected values based on the model already transformed using exp() (see documentation)
fit_data <- fit_data %>% 
  mutate(
    expectation = model$fitted.values,
    threshold = qpois(___,____)
    )

# based on the model make predictions for the last data point which we want to do signal detection for
pred <- predict.glm(model, newdata = pred_data)

pred_data <- pred_data %>%
  mutate(
    expectation = exp(pred),
    threshold = qpois(____,____),
    alarms = pred_data$cases > ______
    )

# combine the two data sets back together
data_with_signals <- bind_rows(fit_data, pred_data)

```

Please plot the result and discuss what you see.

```{r, plot_improved_method-2.1}


```

# Section 3: Optional Task 

## Task 3.1 
We want to do signal detection on another time series with different characteristics. Please plot the time series and describe it.

```{r, ts2-3.1}
# read in the time series 2
ts2 <- readRDS("data/sim2.Rds")

# Plot the time series 2

```

## Task 3.2

Think about a suitable model you would use to do outbreak detection for this time series and fill the gaps. 
```{r, ts2_signal_detection_3.2}
# first we make a fresh environment
rm(fit_data, pred_data, model, data_with_signals, envir = .GlobalEnv)

# split the data set
# data points we use for model fitting, i.e. we build our expectation on are all except the last one
fit_data <- ts2 %>% head(-1)
# the last data point we want to do signal detection for
pred_data <- ts2 %>% tail(1)

# fit the model
model <- glm(formula = cases ~ _______, data = fit_data, family = poisson())
# we generate an expectation for visualisation purposes for the data used for fitting
# model$fitted.values give the expected values based on the model already transformed using exp() (see documentation)
fit_data <- fit_data %>% 
  mutate(
    expectation = model$fitted.values,
    threshold = qpois(_____,_____)
    )

# based on the model make predictions for the last data point which we want to do signal detection for
pred <- predict.glm(model, newdata = pred_data)

pred_data <- pred_data %>%
  mutate(
    expectation = exp(pred),
    threshold = qpois(0.95,expectation),
    alarms = pred_data$cases > ______
    )

# combine the two data sets back together
data_with_signals <- bind_rows(fit_data, pred_data)

```

## Task 3.3

Plot your results and describe what you observe.
```{r, ts2_signal_detection_plot_3.3}



```


# Section 4: Investigate the ESTEC outbreak in Germany 2025 using the SignalDetectionTool

You are the epidemiologist who is in charge of investigating whether there is an ESTEC outbreak going on in Germany and in which subgroups of the population this outbreak is taking place. Apply the SignalDetectionTool to the ESTEC Dataset from Germany in 2025 and find out more about case numbers and which subgroups are affected. Play around with different settings and try to understand the results which are shown in the app.

If you were not able to install the SignalDetectionTool you can use our generated HTML reports instead. 
